---
title: "Use of the DepInfeR package to infer sample-specific protein dependencies from drug-protein profiling and ex-vivo drug response data"
author: "Alina Batzilla, Junyan Lu"
package: DepInfeR
output: 
    BiocStyle::html_document
vignette: >
  %\VignetteIndexEntry{DepInfeR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r message=FALSE, warning=FALSE}
set.seed(123)
library(DepInfeR)
library(tidyr)
library(tidyverse)
library(missForest)
library(parallel)
library(doParallel)
library(glmnet)
library(rlist)
library(matrixStats)
library(dplyr)
library(pheatmap)
library(RColorBrewer)
library(ggrepel)
library(ggbeeswarm)
```

***

# Introduction

DepInfeR is an R package for a computational method that integrates two experimentally accessible input data matrices: the drug sensitivity profiles of cancer cell lines or primary tumors ex-vivo (X), and the drug affinities of a set of proteins (Y), to infer a matrix of molecular protein dependencies of the cancers (ß). DepInfeR deconvolves the protein inhibition effect on the viability phenotype by using regularized multivariate linear regression. It assigns an “dependence coefficient” to each protein and each sample, and therefore could be used to gain causal and accurate understanding of functional consequences of genomic aberrations in a heterogeneous disease, as well as to guide the choice of pharmacological intervention for a specific cancer type, sub-type, or an individual patient.  

This document provide a walk-through of using the **DepInfeR** package to infer sample-specific  protein dependencies from drug-protein affinity profiling and ex-vivo drug response data.


# Data input

As inputs, DepInfeR requires two types of data:
1) a drug-protein affinity profiling dataset containing affinity values between a set of drugs and a set of proteins, and
2) a drug response dataset containing phenotypic data (e.g. viability measurements) of samples in response to a set of drugs.

In this exemplary walk-through analysis we use 
1) drug-protein affinity data from Klaeger et al. 2017, which can be found the supplementary file of the paper (Table_S1 & Table_S2):
https://science.sciencemag.org/content/358/6367/eaan4368/tab-figures-data, and
2) drug response data from the Genomics of Drug Sensitivity in Cancer (GDSC) cancer cell line screening dataset: https://www.cancerrxgene.org/.

A subset of leukemia and breast cancer cell lines was chosen for this analysis. The analyzed cancer types were.

- Diffuse Large B-Cell Lymphoma (DLBC)
- Acute lymphocytic leukemia (ALL)
- Acute myeloid leukemia (AML)
- Breast carcinoma (BRCAHer+ / BRCAHer-)

The Her2 status of the breast cancer cell lines were annotated manually.

As a genetic background of the cancer cell lines, they were annotated with their mutational background. 
The mutational background of the cell lines was retrieved from https://www.cancerrxgene.org/downloads/genetic_features.


This analysis starts with the two data tables that contain the common drugs in both drug-target affinity and drug sensitivity datasets.
```{r}
data(targetsGDSC, drug_response_GDSC)
```


A glance at the drug-target affinity table
```{r}
head(targetsGDSC)
```

A glance at the drug sensitivity table
```{r}
head(drug_response_GDSC)
```

***



# Pre-pocessing the drug-protein dataset

We first need to do some pre-processing of both datasets, to turn them into numeric matrices that can be used as inputs for DepInfeR.

Rename BCR to BCR/ABL to avoid confusion with B-cell receptor (BCR)
```{r}
targetsGDSC <- mutate(targetsGDSC, targetName = ifelse(targetName %in% "BCR", "BCR/ABL", targetName))
```

Turn target table into drug-protein affinity matrix
```{r}
targetMatrix <- dplyr::filter(targetsGDSC, targetClassification == "High confidence") %>% 
    dplyr::select(drugID, targetName, Kd) %>% 
    spread(key = "targetName", value = "Kd") %>%
    remove_rownames() %>% column_to_rownames("drugID") %>% as.matrix()
```


We provide a function, *ProcessTargetResults*, for the pre-processing of the drug-protein affinity matrix with Kd values (or optionally other affinity measurement values at roughly normal distribution). This function can perform the following steps:

- log-transform Kd values (KdAsInput = TRUE)
- arctan-transform log(Kd) values (KdAsInput = TRUE)
- check target similarity and remove highly correlated targets (removeCorrelated = TRUE)

All steps within this function are optional depending on input data. The transformation steps should be performed if the affinity matrix consists of Kd values. If there are highly correlated features within the affinity matrix, they can be removed using the provided function.  The users can also use their own method to preprocess the input drug-target affinity matrix, as long as the distributions of the affinity values is roughly normal and do not have too many highly correlated features.
```{r}
ProcessTargetResults <- processTarget(targetMatrix, KdAsInput = TRUE , removeCorrelated = TRUE)
```

# Preparation of the drug response matrix

## Prepare drug response matrix using the z-score 

The z-score was chosen as a suitable measurement value for our drug screening response matrix as it corresponds to a normalization for each drug over all cell lines. When working with AUC or IC50 values, a suitable normalization of the values is recommended.
In this analysis we used the z-score of the AUC values as the drug response input for DepInfeR.

```{r}
responseMatrix <- filter(drug_response_GDSC, `Drug Id` %in% targetsGDSC$drugID) %>%
  dplyr::select(Drug, `Drug Id`, `Cell line name` , `Cosmic sample Id`, `TCGA classification`, IC50, AUC, `Max conc`,RMSE, `Z score`) %>% 
  dplyr::select(`Drug Id`, `Cell line name`, AUC) %>% 
  tidyr::spread(key = `Cell line name`, value = AUC) %>%
  remove_rownames() %>% column_to_rownames("Drug Id") %>%
  as.matrix()
```

## Assessment of missing values

Currently, DepInfeR does not support input data with missing values. Therefore, the missing values in the input datasets need to be properly handled. The entries with missing values can either be removed or imputed.   

We firstly check the distribution of our missing values across all cell lines.
```{r}
missTab <- data.frame(NA_cutoff = character(0), remain_celllines = character(0), stringsAsFactors = FALSE)
      
for (i in 0 : 138) {
  a <- dim(responseMatrix[,colSums(is.na(responseMatrix)) <= i])[2]
  missTab [i,] <- c(i, a)
}
plot(missTab, type = "l")
```
By looking at the missing value distribution, we choose cell lines with a maximum of 24 missing values per cell line (= 35%) and will imput the remaining missing values by using the MissForest imputation method.

## Subset for cell lines with less than 24 missing values (based on assessment above)
```{r}
responseMatrix <- responseMatrix[,colSums(is.na(responseMatrix)) <= 24]
```

## MissForest imputation
```{r}
impRes <- missForest(t(responseMatrix))
imp_missforest <- impRes$ximp

responseMatrix_imputed <- t(imp_missforest)
```

## Calculate column-wise z-score
```{r}
responseMatrix_scaled <- t(scale(t(responseMatrix_imputed)))
```

# Combine the feature and reponse matrix for regression model

In the step, we will also subset the drug-target affinity matrix and the drug response matrix to only keep the drugs present in both matrix. 
```{r}
targetInput <- ProcessTargetResults$targetMatrix

overlappedDrugs <- intersect(rownames(responseMatrix_scaled), rownames(targetInput))

targetInput <- targetInput[overlappedDrugs,]
responseInput <- responseMatrix_scaled[overlappedDrugs,]
```

***
# Multivariant model for protein dependence prediction

## Multi-target LASSO model

Perform multivariant LASSO regression based on a drug-protein affinity matrix and a drug response matrix.   
In this walk-through we only use 20 repetitions to save time.  In a real application, the number of repetitions can be larger, e.g. repeats = 100, to gain more robust estimations.
```{r}
set.seed(333)
result <- runLASSOregression(TargetMatrix = targetInput, ResponseMatrix  = responseInput, repeats = 3)
```

Remove targets that were never selected
```{r}
useTar <- rowSums(result$coefMat) != 0
result$coefMat <- result$coefMat[useTar,]
```

Number of selected targets
```{r}
nrow(result$coefMat)
```


# Examples of how to inteprete and perform down-stream analyses on the inferred protein dependence matrix

## Heatmap of protein dependence coefficients

The protein dependence matrix can be nicely visualized in a heatmap. High positive coefficients imply strong reliance of a certain sample on this protein for survival. Proteins with coefficients close to zero are less essential for the survival of cells. Negative coefficients indicate that the viability phenotype benefits from inhibition of the protein. 

```{r GDSC_coefHeatmap, fig.height=12, fig.width=18}
#firstly, we need to load the processed mutation annotation of the cell lines
data("mutation_GDSC")
annoColor <- list(H2O2 = c(`-1` = "red", `0` = "black", `1` = "green"),
                  IL.1 = c(`-1` = "red", `0` = "black", `1` = "green"),
                  JAK.STAT = c(`-1` = "red", `0` = "black", `1` = "green"),
                  MAPK.only = c(`-1` = "red", `0` = "black", `1` = "green"),
                  MAPK.PI3K = c(`-1` = "red", `0` = "black"),
                  TLR = c( `-1` = "red", `0` = "black", `1` = "green"),
                  Wnt = c(`-1` = "red", `0` = "black", `1` = "green"),
                  VEGF = c(`-1` = "red", `0` = "black", `1` = "green"),
                  PI3K.only = c(`-1` = "red", `0` = "black", `1` = "green"),
                  TCGA.classification = c(ALL="#BC3C29FF",LAML="#E18727FF",DLBC="#20854EFF","BRCAHer-"="#0072B5FF",'BRCAHer+'="#7876B1FF"),
                  ARID1A_mut = c(`1` = "black",`0` = "grey80"),
                  EP300_mut = c(`1` = "black",`0` = "grey80"),
                  PTEN_mut = c(`1` = "black",`0` = "grey80"),
                  TP53_mut = c(`1` = "black",`0` = "grey80"),
                  PIK3CA_mut = c(`1` = "black",`0` = "grey80"),
                  BRCA2_mut = c(`1` = "black",`0` = "grey80"),
                  BRCA1_mut = c(`1` = "black",`0` = "grey80"),
                  CDH1_mut = c(`1` = "black",`0` = "grey80"),
                  FBXW7_mut = c(`1` = "black",`0` = "grey80"),
                  NRAS_mut = c(`1` = "black",`0` = "grey80"),
                  ASXL1_mut = c(`1` = "black",`0` = "grey80"),
                  MLL2_mut = c(`1` = "black",`0` = "grey80"),
                  ABL1_trans = c(`1` = "black",`0` = "grey80"),
                  missing_value_perc = c(`0` = "white",`25` = "red")
)
plotTab <- result$coefMat
#Row normalization while keeping sign
plotTab_scaled <- scale(t(plotTab), center = FALSE, scale= TRUE)
plotTab <- t(plotTab_scaled)
levels(mutation_GDSC$TCGA.classification) <- c(levels(mutation_GDSC$TCGA.classification), "BRCAHer-")
mutation_GDSC$TCGA.classification[mutation_GDSC$TCGA.classification=="BRCA"] <- "BRCAHer-"
mutation_GDSC$TCGA.classification <- factor(mutation_GDSC$TCGA.classification, levels = c("ALL", "LAML", "DLBC", "BRCAHer-", "BRCAHer+"))
pheatmap(plotTab, 
         color=colorRampPalette(rev(brewer.pal(n = 7, name = "RdBu")), bias= 1.8)(100),
         annotation_col = mutation_GDSC,
         annotation_colors = annoColor,
         clustering_method = "ward.D2", scale = "none", 
         show_colnames = TRUE, main = "", fontsize = 9, fontsize_row = 10, fontsize_col = 7)
```

## Differential dependence on proteins associated with cancer types and genotypes
In this part, we will test whether the sample-specific protein dependence is associated with cancer types or certain genotypes of those samples.   


Prepare genomic background table
```{r gdsc genomic}
cell_anno_final <- mutation_GDSC %>%
  dplyr::select(-missing_value_perc) %>% 
  dplyr::rename(cancer_type = TCGA.classification) %>% 
  dplyr::filter(rownames(mutation_GDSC) %in% colnames(result$coefMat))
colnames(cell_anno_final) <- str_remove_all(colnames(cell_anno_final),"_mut")
colnames(cell_anno_final) <- str_replace_all(colnames(cell_anno_final),"_trans","_translocation")
```

A customised function to perform association test.
```{r}
diffImportance <- function(coefMat, Annotation) {
  #process genetic background table
  geneBack <- Annotation
  geneBack <- geneBack[colnames(coefMat),]
  keepCols <- apply(geneBack,2, function(x) length(unique(na.omit(x))) >=2 & all(table(x)>6))
  geneBack <- geneBack[,keepCols]

  pTab <- lapply(rownames(coefMat), function(targetName) {
    lapply(colnames(geneBack), function(mutName) {
      impVec <- coefMat[targetName, ]
      genoVec <- geneBack[, mutName]
      resTab <- data.frame(targetName = targetName, mutName = mutName,
                           stringsAsFactors = FALSE)
      if (length(unique(na.omit(genoVec))) == 2) {
        #binary feature, usting t.test
        res <- t.test(impVec ~ genoVec, var.equal = TRUE, na.action = na.omit)
        resTab$p <- res$p.value
        resTab$FC <- (res$estimate[[2]]-res$estimate[[1]])/abs(res$estimate[[1]])
      } else if (length(unique(na.omit(genoVec))) >=3) {
        #using anova
        res <- anova(lm(impVec ~ genoVec, na.action = na.omit))
        #get the group mean difference
        diffTab <- data.frame(val = impVec, gr = genoVec) %>%
          dplyr::filter(!is.na(gr)) %>% dplyr::group_by(gr) %>%
          dplyr::summarise(meanVal = mean(val))
        resTab$p = res$`Pr(>F)`[1]
        resTab$FC = max(diffTab$meanVal)-min(diffTab$meanVal)
      }
      resTab
    }) %>% dplyr::bind_rows()
  }) %>% dplyr::bind_rows() %>% dplyr::arrange(p) %>% dplyr::mutate(p.adj = p.adjust(p, method = "BH"))
  pTab
}
```


Association test between protein dependence and cancer type or mutational background
```{r gdsc t-test}
testRes <-  diffImportance(result$coefMat, cell_anno_final)
```

## Visualize protein associations with cancer type
```{r GDSC_diffCoef_box,fig.height=7, fig.width=16}
CancerType <- testRes %>% dplyr::filter(mutName == "cancer_type") %>% dplyr::filter(p.adj < 0.05, FC > 0.1)
plotTab <- t(scale(t(result$coefMat))) %>% data.frame() %>%
  rownames_to_column("target") %>% gather(key = "CellLine", value = "coef",-target) %>% mutate(Cancer_Type = mutation_GDSC[CellLine,]$TCGA.classification) %>%
  group_by(target, Cancer_Type) %>% mutate(meanCoef = mean(coef)) %>% arrange(meanCoef) %>% ungroup() %>%
  mutate(target = factor(target, levels = unique(target))) 
plotTab <- plotTab %>% dplyr::filter(target %in% CancerType$targetName)
plotTab$Cancer_Type <- factor(plotTab$Cancer_Type, levels = c("ALL", "LAML","DLBC", "BRCAHer-", "BRCAHer+"))
ggplot(plotTab, aes(x = target, y = coef, group=Cancer_Type)) +
 geom_jitter(
  aes(color = Cancer_Type), 
  position = position_jitterdodge(jitter.width = 0.2, dodge.width = 0.8),
  size = 1.2
  ) +
  stat_summary(
    fun= mean, fun.min=mean, fun.max=mean, colour="grey25",
    geom = "crossbar",  size = 0.8,
    position = position_dodge(0.8)
    ) +
  scale_color_manual(values= c("#BC3C29FF","#E18727FF","#20854EFF","#0072B5FF","#7876B1FF"),
                     guide = guide_legend(override.aes = list(size = 3) )) +
  ggtitle("Protein dependence associated with cancer type") + ylab("Protein dependence coefficient") + xlab("Protein") + theme_bw() + 
  geom_vline(xintercept =seq(from = 1.5, to = 8.5, by = 1), color="darkgrey") + labs(color = "Cancer Type")
```

## Visualize P-values of significant associations between protein dependence and mutational background
```{r gdsc_pval_vis, fig.height=6, fig.width=10}
colList2 <- c(`not significant` = "grey80", mutated = "#BC3C29FF", unmutated = "#0072B5FF")
pos = position_jitter(width = 0.15, seed = 10)
plotTab <- testRes %>% dplyr::filter(mutName != "cancer_type") %>% mutate(type = ifelse(p.adj > 0.1, "not significant", 
                                          ifelse(FC >0, "mutated","unmutated"))) %>%
  mutate(varName = ifelse(type == "not significant","",targetName)) %>%
  mutate(p.adj = ifelse(p.adj <1e-5, 1e-5,p.adj))
#subset for mutation with at least one significant associations
plotMut <- unique(filter(testRes, p.adj <= 0.1)$mutName)
plotTab <- plotTab %>% dplyr::filter(mutName %in% plotMut) 
plotTab$type <- factor(plotTab$type, levels = c("mutated", "unmutated", "not significant"))
p <- ggplot(data=plotTab, aes(x= mutName, y=-log10(p.adj),
                              col=type, label = varName))+ 
  geom_text_repel(position = pos, color = "black", size= 6, force = 3) +
  geom_hline(yintercept = -log10(0.1), linetype="dotted", color = "grey20") + 
  geom_point(size=3, position = pos) + 
  ylab(expression(-log[10]*'('*adjusted~italic("P")~value*')')) + xlab("Mutation") +
  scale_color_manual(values = colList2) + 
  scale_y_continuous(trans = "exp", limits = c(0,2.5), breaks = c(0,1,1.5,2)) + 
  #annotate(geom = "text", x = 0.5, y = -log10(0.1) - 0.25, label = "10% FDR", size=7, col = "grey20") +
  coord_flip()  + labs(col = "Higher dependence in")  +
  theme_bw() +
  theme(legend.position = c(0.80,0.2),
        legend.background = element_rect(fill = NA),
        legend.text = element_text(size=14),
        legend.title = element_text(size=16),
        axis.title = element_text(size=18),
        axis.text = element_text(size=18))
plot(p)
#ggsave("test.pdf",height = 4, width = 8)
```

## Boxplot visualization

Function for boxplot, given a t-test result table, coef/frequency matrix and cell line annotation object
```{r}
#Function to format floats
formatNum <- function(i, limit = 0.01, digits =1, format="e") {
  r <- sapply(i, function(n) {
    if (n < limit) {
      formatC(n, digits = digits, format = format)
    } else {
      format(n, digits = digits)
    }
  })
  return(r)
}

plotDiffBox <- function(pTab, coefMat, cellAnno, fdrCut = 0.05) {
  #filter
  pTab.sig <- filter(pTab, p.adj <= fdrCut)

  #process genetic background table
  geneBack <- cellAnno
  geneBack <- geneBack[colnames(coefMat),]

  pList <- lapply(seq(nrow(pTab.sig)), function(i) {
    geno <- pTab.sig[i,]$mutName
    target <- pTab.sig[i,]$targetName
    pval <- pTab.sig[i,]$p
    plotTab <- tibble(id = colnames(coefMat),
                      mut = geneBack[, geno],
                      val = coefMat[target,]) %>%
      filter(!is.na(mut))

    if (str_detect(geno, "trisomy12|gain|del")) {
      plotTab <- mutate(plotTab, mut = ifelse(mut %in% c("1",1), "present","absent")) %>%
        mutate(mut = factor(mut, levels = c("absent", "present")))
    } else {
      plotTab <- mutate(plotTab, mut = ifelse(mut %in% c("M","1",1), "Mutated","Unmutated")) %>%
        mutate(mut = factor(mut, levels = c("Unmutated","Mutated")))
    }

    #count cases
    numTab <- group_by(plotTab, mut) %>%
        summarise(n=length(id))

    plotTab <- left_join(plotTab, numTab, by = "mut") %>%
      mutate(mutNum = sprintf("%s\n(n=%s)", mut,n)) %>%
      arrange(mut) %>%
      mutate(mutNum = factor(mutNum, levels = unique(mutNum)))

    if (geno == "FLT3.ITD") {
      genoType = "mutation"
      geno <- "FLT3-ITD"
    } else if (str_detect(geno, "trisomy12|gain|del|IGHV.status")) {
      genoType = ""
      geno <- str_replace(geno, "[.]"," ")
    } else {
      genoType = "mutations"
    }

    titleText <- sprintf("%s ~ %s %s", target, geno, genoType)
    pval <- formatNum(pval, digits = 1, format="e")
    titleText <- bquote(atop(.(titleText), "("~italic("P")~"="~.(pval)~")"))

    ggplot(plotTab, aes(x = mutNum,y = val)) +
      stat_boxplot(geom = "errorbar", width = 0.3) +
      geom_boxplot(outlier.shape = NA, col="black", width=0.4) +
      geom_beeswarm(cex=2, size =2, aes(col = mutNum)) + theme_classic() +
      xlab("") + ylab("Protein dependence") + ggtitle(titleText) + xlab("") +
      scale_color_manual(values = c("#0072B5FF","#BC3C29FF")) +
      theme(axis.line.x = element_blank(), axis.ticks.x = element_blank(),
            axis.title = element_text(size=18),
            axis.text = element_text(size=18),
            plot.title = element_text(size= 20, face = "bold", hjust = 0.5),
            legend.position = "none")
  })
  names(pList) <- paste0(pTab.sig$targetName,"_", pTab.sig$mutName)
  return(pList)

}
```


Visualization of exemplary association between NRAS mutation status and MAP2K2 dependence visualized in a beeswarm plot
```{r gdsc boxplot, fig.height=4.5, fig.width=5}
pList <- plotDiffBox(testRes, result$coefMat, cell_anno_final, 0.05)
pList$MAP2K2_NRAS
```

#Session info
```{r}
sessionInfo()
```

